Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-01 paper/metamonads.tex
Update the introduction.  Almost done first draft of the intro.

\documentclass[11pt]{llncs}

@@

\usepackage{hyphenat}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{amstext}
\usepackage{url}

@@

@@

\newcommand{\omitnow}[1]{}
\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}

@@

Our goal was to implement a completely modular, textbook-style 
modular design \cite{journals/cacm/parnas72a}, yet 
have our code be very efficient.  Gaussian elimination, as it is 
well-known, already contains many different aspects \cite{carette04},
was chosen as a representative example.

@@

While many people still make do with string-based code generation,
this is not much different than writing code in a dynamically typed
interpreted scripting language.  It works, it is convenient, but it
does not scale well, it is a debugging nightmare, and usually suffers
most from Software Aging \cite{parnas_aging}.  As with other tasks in
programming, the answer is to turn to statically typed languages,
which provide strong guarantees about the soundness of the resulting
code.  MetaOCaml (cite website?) \cite{CTHL03} in particular stands
out: not only it guarantees the absense of problems such as
accidental variable capture -- so-called hygiene cite? \oleg{should we
  also mention automatic well-formedness}, MetaOcaml also
statically ensures that both the code generator and the generated code
are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
programming in C++ and Template Haskell do offer some amount of static
typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml\footnote{and
  MetaML.}  offers stronger guarantees.  As well, since MetaOCaml is
implemented on top of the very solid Ocaml language, one accrues
additional benefits.

@@

In MetaOCaml, generated code cannot be examined and optimized at all. It must
be generated just right.

- The FFT paper
showed one way: abstract interpretation. But more problems remain
(cite Padua et al paper at Haskell Workshop04): generating names in
the presence of `if', 'while' etc. control operators (simple problem
of generating names solved in the FFT paper); making CPS code clear
(note that many authors are afraid of CPS code because it quickly
becomes unreadable; we need CPS for name generation); compositionality
of various (often interleaving: explain: determinant computation
requires attention at several places) aspects and expressing
dependencies among them; expressing more static guarantees about the
generated code (e.g., not attempting full (fractional) division for
the integer domain; generating full-pivoting code for the rational
domain without |better_than| operation). In short: can we achieve all
of the abstractions needed for the flexible Gaussian Elimination? Can
we eliminate all of that abstraction's overhead in the generated code
(without using intensional code analysis -- that is, maintaining the
guarantees of MetaOCaml and using pure generative approach).

@@

\oleg{explictly enumerate our contributions, not foiregtting the mdo
  notation. The subsubsection that outlines what the other sections
  are for.}

>>>>

\documentclass{llncs}

@@

\usepackage{hyphenat}
\usepackage{comment}
\usepackage{amsmath,amssymb}
\usepackage{amstext}
\usepackage{url}

@@

% \usepackage[backref,colorlinks,bookmarks=true]{hyperref}

@@

\newcommand{\omitnow}[1]{}
\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}

@@

Our goal was to implement a completely modular, textbook-style 
modular design \cite{journals/cacm/parnas72a}, yet 
have our code be very efficient.  Gaussian elimination, as it is a  
well-known algorithm, already contains many different aspects 
\cite{carette04}, was chosen as a representative example.

@@

While many people still make do with string-based code generation,
this is not much different than writing code in a dynamically typed
interpreted scripting language.  It works, it is convenient, but it
does not scale well, it is a debugging nightmare, and usually suffers
most from Software Aging \cite{parnas_aging}.  As with other tasks in
programming, the answer is to turn to statically typed languages,
which provide strong guarantees about the soundness of the resulting
code.  MetaOCaml \cite{CTHL03,metaocaml-org} in particular stands
out: not only it guarantees well-formedness and the absense of problems 
such as accidental variable capture \cite{HygienicMacros}
\jacques{is this the citation you meant?}
MetaOcaml also
statically ensures that both the code generator and the generated code
are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
programming in C++ and Template Haskell do offer some amount of static
typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml\footnote{and
  MetaML.} offers stronger guarantees.  As well, since MetaOCaml is
implemented on top of the very solid Ocaml language, one accrues
additional benefits.

@@

In MetaOCaml, generated code cannot be examined and optimized at all. It must
be generated just right.  For simple code, this is relatively easy
(see \cite{TahaThesis} for many examples).  For more complex examples,
new techniques are necessary.  \cite{KiselyovTaha} introduces one
such tool: abstract interpretation.  But more problems remain:
\jacques{can't find this 'Padua' paper at the 2004 Haskell Workshop 
you refer to}: 
\begin{itemize}
	\item Generating names in the presence of ``if'', ``while'', and other
		control operators. \cite{KiselyovTaha} showed how to solve
		this problem in simple cases.
	\item Making continuation-passing style (CPS) code clear.  Many
		authors understandably shy away from CPS code as it quickly
		becomes unreadable.  But this is needed for proper name
		generation.
	\item Compositionality of various interleaving aspects, as well
		as expressing dependencies among them.  For example, if we
		choose to compute the determinant of a matrix while performing
		Gaussian Elimination, this requires inserting code in 
		several specific places in the algorithm.  The method to
		compute this determinant crucially depends on whether 
		a division-based or a fraction-free Gaussian Elimination is
		used.
	\item As mentionned above, there are dependencies between
		certain algorithmic choices and properties of the
		underlying domain, and we would like to statically guarantee
		that inappropriate combinations cannot be chosen.  For instance,
		one should not attempt to use full division when dealing
		with matrices of exact integers, nor is it worthwhile to use
		full pivoting on a matrix of rational integers ($\mathbb Q $).
\end{itemize}
The question we want to answer is:
Can we achieve all of the abstractions needed for a flexible 
Gaussian Elimination, and can
we eliminate all of that abstraction's overhead in the generated code?
Of course, we want to do this without using
intensional code analysis -- that is, maintaining the
guarantees of MetaOCaml and using a purely generative approach.

@@

This work makes the following contributions:
\begin{itemize}
	\item mdo notation
	\item use of monads for meta-programming
	\item use of generative functors
	\item use of type constraints to ensure correctness
\end{itemize}
\oleg{The subsubsection that outlines what the other sections
  are for.}
